Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-06-23 paper/paper1/ms.tex
Some formatting-- figure placement in two column mode

\documentclass[modern]{aastex631}
\usepackage{hyperref}
\bibliographystyle{aasjournal}
%\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}
\usepackage{outlines}
\usepackage{amsmath,bm}
\definecolor{belize}{RGB}{41, 128, 185}
\definecolor{peter}{RGB}{52, 152, 219}
\definecolor{nephritis}{RGB}{39, 174, 96}
\definecolor{asbestos}{RGB}{127, 140, 141}
\hypersetup{linkcolor=asbestos, 
citecolor=belize,filecolor=asbestos,urlcolor=peter}

@@

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.45\textwidth]{figures/N_lines_vs_Teff_prom.png}
    \includegraphics[width=0.45\textwidth]{figures/N_lines_vs_Teff_prom.png}
    %\includegraphics[width=0.45\textwidth]{figures/line_density_demo.png}
    \caption{Scaling of prominence and density of spectral lines.
        \emph{Left:} Number of lines versus effective temperature for PHOENIX models truncated to IGRINS (\textbf{blue, connected} points) and HPF (\emph{red, free-standing} points) bandwidths, for different prominence thresholds of $0.02$, $0.01$, and $0.005$.
        \textbf{Right:} Number of lines per 100 \AA\ wavelength bin for stellar (\emph{blue, upper envelope of steps}) and telluric (\emph{orange, lower envelope of steps}), illustrated for a $T_\mathrm{eff}=4700$~K, $\log{g}=4.5$ PHOENIX model and a $T=290 \mathrm{K}$, relative humidity of 40\% TelFit model.}
    \label{fig_Nlines_vs_teff}
\end{figure}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.98\textwidth]{figures/blase_flowchart01.png}
    \caption{Visual flowchart of \texttt{blas\'e}.  The technique starts with a precomputed synthetic stellar spectrum and a precomputed synthetic telluric spectrum with properties close to those previously estimated for the target.  Both the stellar and telluric spectra get cloned.  The stellar model is warped to its extrinsic properties, and then the stellar and telluric models get multiplied together.  Next, this joint model is convolved with an instrumental kernel and resampled to the data spectrum.  Finally, the joint model parameters are optimized to match the data, subject to regularization (not shown).  The final products are semi-empirical super-resolution stellar and telluric templates, the $RV$ and $v\sin{i}$ of the star, and the line amplitude, location, width, and shape of all 10,000+ stellar and telluric lines. \authorcomment1{Some of the spectra are placeholders until we re-run the end-to-end-fitting.  Feedback on this figure layout is sought.} }
    \label{blase_flowchart}
\end{figure}

@@

\begin{equation} \bm{\bar{F}} =
    \begin{pmatrix}
        \cdots \includegraphics[height=1cm]{figures/voigt_01.pdf}  \cdots & \\
        \cdots \includegraphics[height=1cm]{figures/voigt_02.pdf}  \cdots & \\
        \vdots                                                            & \\
        \cdots \includegraphics[height=1cm]{figures/voigt_03.pdf} \cdots  &
    \end{pmatrix} \mapsto
    \begin{pmatrix}
        \includegraphics[height=1cm]{figures/voigt_01b.pdf} & \\
        \includegraphics[height=1cm]{figures/voigt_02b.pdf} & \\
        \vdots                                              & \\
        \includegraphics[height=1cm]{figures/voigt_03b.pdf} &
    \end{pmatrix} = \bm{\hat{F}}
\end{equation}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_clone_demo.png}
    \caption{PHOENIX spectrum cloned with blas\'e.  This $T_{\mathrm{eff}}=4700\;$K,  $\log{g}=4.5$ solar metallicity model has 9,028 individual cloned spectral lines, each with 3 tuned parameters.  The pictured 50 \AA~ chunk contains 121 spectral lines and represents about 1$\%$ of the entire spectral bandwidth that was cloned.  Some flaws can be seen near the cores of deep lines, or wings of broad lines.}
    \label{fig_cloned_spectrum_demo}
\end{figure}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_clone_demo_zoom.png}
    \caption{Zoom in of the region between the vertical gray bars in Figure \ref{fig_cloned_spectrum_demo}. The cloned model has 7 spectral lines describing the 400 pixels in this 4 \AA~ chunk.  The native PHOENIX pixel sampling can be see as the boxy steps in both the native and cloned model.  }
    \label{fig_zoom_cloning_performance}
\end{figure}

@@

\begin{eqnarray}
    \mathrm{regularization} \equiv \ln{\mathrm{prior}}  = \frac{1}{2} \sum_{j=1}^{N_\mathrm{lines}} \Big(\frac{\ln{a_j} - \ln{a^\prime_j }}{\Lambda}\Big)^2
\end{eqnarray}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_WASP69_demo.png}
    \caption{WASP 69 observed with HPF, compared to a $T_{\mathrm{eff}}=4700\;$K,  $\log{g}=4.5$ solar metallicity PHOENIX model warped to $v\sin{i}=2.2$~km/s, $RV=-9.6$~km/s, and HPF resolving power.}
    \label{fig_WASP69_demo}
\end{figure}

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_WASP69_regularized.png}
    \caption{Semi-empirical model of WASP 69 transfer-learned with \texttt{blas\'e}, employing a regularization prior on the learned amplitudes.}
    \label{fig_WASP69_transferred}
\end{figure}

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_super_resolution_template.png}
    \caption{The native-resolution semi-empirical model transfer-learned from WASP 69 HPF data. The revised model can be viewed as a super-resolution deconvolution of the HPF spectrum.}
    \label{fig_WASP69_regularized}
\end{figure}

@@

Currently, each training epoch ``sees'' the entire dataset, a setup dubbed ``full-batch'' gradient descent.  An alternative scheme allows training with only a portion of the entire dataset at a time in ``minibatches''.  The massive data volumes in modern Neural Network applications cannot fit into the GPU memory, so minibatches are a necessity.  Our meager $1$ MB dataset can easily fit into the GPU memory, but our model can be large if we have a large number of pixels or lines or both.  So while minibatches may not be required due to data size limitations they may be useful for particularly large models.  Minibatches also act as a form of regularization, the principal source of stochasticity in the Stochastic Gradient Descent algorithm, which tends to have better convergence than full-batch Gradient Descent \citep{2016arXiv160904747R}.

@@

It is much easier to reverse-engineer longitudinally asymmetric features than longitudinally symmetric ones.  The latter requires exact knowledge of the underlying spectral template.  Isolated deep spectral lines constitute the only practical scenario where exact knowledge can be claimed.  Isolated spectral lines may be scarce or absent for M-dwarfs and brown dwarfs where lines blend ostensibly in an inseparable way, counfounding Doppler imaging.

@@

\subsection{Lineshapes choice and tradeoffs}
\begin{outline}
    \1 Faced with a question: what lineshape to use
    \2 To First order the lines are Gaussian or Lorentzian
    \2 To Second-order the lines are Voigt Profiles
    \2 In detail the lines are the result of radiative transfer that can smear the Voigt profile
    \2 In practice the high-res lines get convolved with instrumental profiles, so details don't matter too much
    \1 Computational tradeoff of Lorentzian/Gaussian versus Pseudo-Voigt versus Voigt
    \2 Additional Implementation challenge: Fadeeva function not implemented in PyTorch
    \1 We choose a pseudo-Voigt as a balance between computation and adequate accuracy
\end{outline}

>>>>

\documentclass[twocolumn]{aastex631}
\usepackage{hyperref}
\bibliographystyle{aasjournal}
%\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}
\usepackage{outlines}
\usepackage{amsmath,bm}
\definecolor{belize}{RGB}{41, 128, 185}
\definecolor{peter}{RGB}{52, 152, 219}
\definecolor{nephritis}{RGB}{39, 174, 96}
\definecolor{asbestos}{RGB}{127, 140, 141}
\hypersetup{linkcolor=asbestos, 
citecolor=belize,filecolor=asbestos,urlcolor=peter}

\hyphenation{long-it-udin-ally hy-po-thet-ical de-mon-stra-tion}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.45\textwidth]{figures/N_lines_vs_Teff_prom.png}
    %\includegraphics[width=0.45\textwidth]{figures/N_lines_vs_Teff_prom.png}
    \includegraphics[width=0.45\textwidth]{figures/line_density_demo.png}
    \caption{Scaling of prominence and density of spectral lines.
        \emph{Left:} Number of lines versus effective temperature for PHOENIX models truncated to IGRINS (\textbf{blue, connected} points) and HPF (\emph{red, free-standing} points) bandwidths, for different prominence thresholds of $0.02$, $0.01$, and $0.005$.
        \textbf{Right:} Number of lines per 100 \AA\ wavelength bin for stellar (\emph{blue, upper envelope of steps}) and telluric (\emph{orange, lower envelope of steps}), illustrated for a $T_\mathrm{eff}=4700$~K, $\log{g}=4.5$ PHOENIX model and a $T=290 \mathrm{K}$, relative humidity of 40\% TelFit model.}
    \label{fig_Nlines_vs_teff}
\end{figure}

@@

@@

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=0.98\textwidth]{figures/blase_flowchart01.png}
    \caption{Visual flowchart of \texttt{blas\'e}.  The technique starts with a precomputed synthetic stellar spectrum and a precomputed synthetic telluric spectrum with properties close to those previously estimated for the target.  Both the stellar and telluric spectra get cloned.  The stellar model is warped to its extrinsic properties, and then the stellar and telluric models get multiplied together.  Next, this joint model is convolved with an instrumental kernel and resampled to the data spectrum.  Finally, the joint model parameters are optimized to match the data, subject to regularization (not shown).  The final products are semi-empirical super-resolution stellar and telluric templates, the $RV$ and $v\sin{i}$ of the star, and the line amplitude, location, width, and shape of all 10,000+ stellar and telluric lines. \authorcomment1{Some of the spectra are placeholders until we re-run the end-to-end-fitting.  Feedback on this figure layout is sought.} }
    \label{blase_flowchart}
\end{figure*}

@@

\begin{equation} 
    \begin{aligned}
    \bm{\bar{F}} &=
    \begin{pmatrix}
        \cdots \includegraphics[height=1cm]{figures/voigt_01.pdf}  \cdots & \\
        \cdots \includegraphics[height=1cm]{figures/voigt_02.pdf}  \cdots & \\
        \vdots                                                            & \\
        \cdots \includegraphics[height=1cm]{figures/voigt_03.pdf} \cdots  &
    \end{pmatrix} \\
    \bm{\bar{F}}&\mapsto \bm{\hat{F}} =\begin{pmatrix}
        \includegraphics[height=1cm]{figures/voigt_01b.pdf} & \\
        \includegraphics[height=1cm]{figures/voigt_02b.pdf} & \\
        \vdots                                              & \\
        \includegraphics[height=1cm]{figures/voigt_03b.pdf} &
    \end{pmatrix}
\end{aligned}
\end{equation}

@@

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_clone_demo.png}
    \caption{PHOENIX spectrum cloned with blas\'e.  This $T_{\mathrm{eff}}=4700\;$K,  $\log{g}=4.5$ solar metallicity model has 9,028 individual cloned spectral lines, each with 3 tuned parameters.  The pictured 50 \AA~ chunk contains 121 spectral lines and represents about 1$\%$ of the entire spectral bandwidth that was cloned.  Some flaws can be seen near the cores of deep lines, or wings of broad lines.}
    \label{fig_cloned_spectrum_demo}
\end{figure*}

@@

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_clone_demo_zoom.png}
    \caption{Zoom in of the region between the vertical gray bars in Figure \ref{fig_cloned_spectrum_demo}. The cloned model has 7 spectral lines describing the 400 pixels in this 4 \AA~ chunk.  The native PHOENIX pixel sampling can be see as the boxy steps in both the native and cloned model.  }
    \label{fig_zoom_cloning_performance}
\end{figure*}

@@

\begin{eqnarray}
    \mathrm{regularization} \equiv -\ln{\mathrm{prior}}  = \frac{1}{2} \sum_{j=1}^{N_\mathrm{lines}} \Big(\frac{\ln{a_j} - \ln{a^\prime_j }}{\Lambda}\Big)^2
\end{eqnarray}

@@

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_WASP69_demo.png}
    \caption{WASP 69 observed with HPF, compared to a $T_{\mathrm{eff}}=4700\;$K,  $\log{g}=4.5$ solar metallicity PHOENIX model warped to $v\sin{i}=2.2$~km/s, $RV=-9.6$~km/s, and HPF resolving power.}
    \label{fig_WASP69_demo}
\end{figure*}

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_WASP69_regularized.png}
    \caption{Semi-empirical model of WASP 69 transfer-learned with \texttt{blas\'e}, employing a regularization prior on the learned amplitudes.}
    \label{fig_WASP69_transferred}
\end{figure*}

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/blase_super_resolution_template.png}
    \caption{The native-resolution semi-empirical model transfer-learned from WASP 69 HPF data. The revised model can be viewed as a super-resolution deconvolution of the HPF spectrum.}
    \label{fig_WASP69_regularized}
\end{figure*}

@@

Currently, each training epoch sees the entire dataset, a setup dubbed \emph{full-batch} gradient descent.  
An alternative scheme allows training with only a portion of the entire dataset at a time in \emph{minibatches}.  The massive data volumes in modern Neural Network applications cannot fit into the GPU memory, so minibatches are a necessity.  Our meager 1 MB dataset can easily fit into the GPU memory, but our model can be large if we have a large number of pixels or lines or both.  
So while minibatches may not be required due to data size limitations they may be useful for particularly large models.  
Minibatches also act as a form of regularization, the principal source of stochasticity in the Stochastic Gradient Descent algorithm, which tends to have better convergence than full-batch Gradient Descent \citep{2016arXiv160904747R}.

@@

It is easier to reverse-engineer longitudinally asymmetric features than longitudinally symmetric ones.  
The latter requires exact knowledge of the underlying spectral template.  Isolated deep spectral lines constitute the only practical scenario where exact knowledge can be claimed.  Isolated spectral lines may be scarce or absent for M-dwarfs and brown dwarfs where lines blend ostensibly in an inseparable way, counfounding Doppler imaging.

@@

\subsection{Lineshapes choice and tradeoffs}
\begin{outline}
    \1 Faced with a question: what lineshape to use
    \2 To First order the lines are Gaussian or Lorentzian
    \2 To Second-order the lines are Voigt Profiles
    \2 In detail the lines are the result of radiative transfer that can smear the Voigt profile
    \2 In practice the high-res lines get convolved with instrumental profiles, so details don't matter too much
    \1 Computational tradeoff of Lorentzian/Gaussian versus Pseudo-Voigt versus Voigt
    \2 Additional challenge: Fadeeva function not in PyTorch
    \1 We choose a pseudo-Voigt as a balance between computation and adequate accuracy
\end{outline}
